Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
automata fixed.
fig 6 fixed
CFG->NFA->DFA

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item For each function $\mathit{f}$, equations defining \Df{f}{i} for
  use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item  For  each  function   $\mathit{f}$  (including  $\mainpgm$  for
  $e_\mainpgm$)  an equation  defining liveness  at each  GC point  of
  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
\begin{eqnarray*}
  \var{\Lanv{1}{\xl}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted  $\mathscr{L}(\var{\Lanv{i}{x}})$, is  the
desired solution  of \Lanv{i}{x}.  However,  note that the  strings in
the  language are  over alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,
\clazy \rbrace$, but  we want {\em forward} access paths  only, i. e.,
access paths over the alphabet $\lbrace \acar, \acdr \rbrace$. In
other words,  the decision  problem that we  are interested  in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification  is  continued  till  a fixed  point  is  reached.  The
algorithm for this appears in~\cite{asati14lgc}, Figure[].  Finally we
replace the  $\clazy$ edges by  $\epsilon$ and check if  the automaton
has  valid path  (consisting  of only  \acar\ and  \acdr)  to a  final
state. If it has a valid path,  we remove the \clazy\ edge and mark the
source node as  final, else we remove the \clazy\  edge without marking
the source  node as final.   Once the  \clazy\ edges are  eliminated we
convert the automaton into a deterministic automaton.

\begin{figure}[t!]
  \begin{tabular}{cc}
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \ncline{t1}{t2}
  \putnode{l0}{t1}{7}{2}{\clazy}
  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
  \putnode{l1}{t1}{0}{10}{\acdr/\clazy}
  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-.5]{t2}{t2}
  \putnode{l2}{t2}{0}{10}{\clazy}
\end{pspicture}
&
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-.5]{t1}{t1}
  \putnode{l1}{t1}{0}{10}{\acdr}
\end{pspicture}
\\
(a) & (b) \\
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \ncline{t1}{t2}
  \putnode{l0}{t1}{5}{2}{\clazy}
  \ncline{t2}{t3}
  \putnode{l1}{t2}{5}{2}{\bcar}
  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
  \putnode{l2}{t3}{0}{10}{\acdr}
\end{pspicture}
&
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(40,12) %\psframe(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\Lanv{9}{\pa}}
  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \putnode{t3}{t1}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \ncline{t1}{t3}
  \putnode{l1}{t1}{5}{2}{\bcar}
  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.5]{t3}{t3}
  \putnode{l2}{t3}{0}{10}{\acdr}
\end{pspicture}
\\
(c) & (d)  \end{tabular}
\caption{(a) The  grammar rules for $\langle  \Df{\length}{1} \rangle$
  converted into an automaton, and simplified in (b). (c) and (d) show
  the   automata  for   \Lanv{9}{\pa}.   }\label{fig:example-automata}
\figrule
\end{figure}

>>>>

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item For each function $\mathit{f}$, equations defining \Df{f}{i} for
  use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item  For  each  function   $\mathit{f}$  (including  $\mainpgm$  for
  $e_\mainpgm$)  an equation  defining liveness  at each  GC point  of
  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
\begin{eqnarray*}
  \var{\Lanv{1}{\xl}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{9}{\pa}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted  $\mathscr{L}(\var{\Lanv{i}{x}})$, is  the
desired solution  of \Lanv{i}{x}.  However,  note that the  strings in
the  language are  over alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,
\clazy \rbrace$, but  we want {\em forward} access paths  only, i. e.,
access paths over the alphabet $\lbrace \acar, \acdr \rbrace$. In
other words,  the decision  problem that we  are interested  in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labelled \clazy, we check if the source node
of the edge has  a valid path to a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labelled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.

\begin{figure}[t!]
  \begin{tabular}{cc}
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(50,12)
  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
  \putnode{t1}{t0}{15}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \putnode{t2}{t1}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \ncline{t1}{t2}
  \putnode{l0}{t1}{7}{2}{\clazy}
  \nccurve[angleA=45, angleB=135, ncurv=4, nodesep=-1]{t1}{t1}
  \putnode{l1}{t1}{0}{10}{\acdr, \clazy}
  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t2}{t2}
  \putnode{l2}{t2}{0}{10}{\clazy}
\end{pspicture}
&
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(30,12)
  \putnode{t0}{origin}{5}{0}{\var{\Df{\length}{1}}}
  \putnode{t1}{t0}{15}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \nccurve[angleA=45, angleB=135, ncurv=3, nodesep=-1]{t1}{t1}
  \putnode{l1}{t1}{0}{10}{\acdr}
\end{pspicture}
\\
(a) & (b) \\
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(50,12) %\psframe(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \putnode{t2}{t1}{12}{0}{\pscirclebox{\mbox{\ \ \ \ }}}
  \putnode{t3}{t2}{12}{0}{\pscirclebox[doubleline=true]{\mbox{\ \ \ \ }}}
  \psset{arrows=->}
  \ncline{t0}{t1}
  \ncline{t1}{t2}
  \putnode{l0}{t1}{5}{2}{\clazy}
  \ncline{t2}{t3}
  \putnode{l1}{t2}{5}{2}{\bcar}
  \nccurve[angleA=55, angleB=125, ncurv=3, nodesep=-.8]{t3}{t3}
  \putnode{l2}{t3}{0}{10}{\acdr}
\end{pspicture}
&
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,-5)(20,12) %\psframe(0,-5)(40,12)
  \putnode{t0}{origin}{5}{0}{\var{\Lanv{9}{\pa}}}
  \putnode{t1}{t0}{9}{0}{\pscirclebox{\mbox{\ \ \ \ }}} \hspace{5mm}
  \psset{arrows=->}
  \ncline{t0}{t1}
\end{pspicture}
\\
(c) & (d)  \end{tabular}
\caption{(a) The  grammar rules for \var{\Df{\length}{1}}
  converted into an automaton, and simplified in (b). (c) and (d) show
  the   same  for   \var{\Lanv{9}{\pa}}.   }\label{fig:example-automata}
\figrule
\end{figure}
